---
title: Mtcars exploratory data analysis
author: "Americo"
date: "28 settembre 2015"
output: html_document
---
The question of interest are:

- Is an automatic or manual transmission better for MPG”?
- Can you quantify the MPG difference between automatic and manual transmissions?

##Mtcars dataset

```{r packages}
require(car); require(dplyr); require(datasets); require(GGally); require(psych); require(tidyr); require(ggplot2); require(lsr)
```

```{r loading mtcars}
?mtcars
str(mtcars)
```

Le variabili sono tutte numeriche; alcune converrà trasformarle in fattore.
Le variabili sono:

-mpg: miglia per gallone;
- cyl: numero di cilindri;
- disp: cilindrata;
- hp: cavalli
- drat: Rear axle ratio; più basso è meno consumi;
- wt: peso in pound diviso mille, quindi se fossimo in italia tonnellate invece di kg
- qsec: tempo per fare da fermo un quarto di miglia(immagino che più breve sia più consumi);
- V/S: alignment of the cylinders - either a V shape or Straight
- am: trasmissione automatica del cambio;
- gear: numero di marce anteriori;
- carb: numero di carburatori

##Exploratory data analysis

###Missing values

```{r looking for NA}
all(complete.cases(mtcars))
```

Nessuna riga ha valori mancanti

###Analisi monovariate

```{r summary}
summary(mtcars)
describe(mtcars)
sapply(mtcars, table)
```

Ci sono alcune variabili come `carb``gear` `am` `vs`e `cyl` che hanno davvero pochi valori, e bisogna valutare se non sia il caso di considerarli qualitative. 

###Scatterplot matrix

```{r matrice scatterplot}
ggpairs(mtcars)
```
##Main Effect

Analyzing the main effect of `am` on `mpg` we see an influence of trasmission on fule consumption:

```{r am 1}
mtcars_agg1  <- mtcars %>%
  group_by(am) %>%
  summarize(mean_mpg = mean(mpg),
            n= n())
mtcars_agg1
```

So you see a difference. Is it significant? Let's perform a *one-way Anova*:

```{r am anova}
am_aov <- aov(mpg ~ am, mtcars)
summary(am_aov)
etaSquared(am_aov)
```

It seems significant and with a tangible effect.

But what if there is some confounding variable? From the scatterplot matrix you see that `am` has a high correlation with `wt`, `drat` and `gear`. So let's explore the relationship between `am` and `mpg` adjusting for this other variables.

###wt
```{r am e wt plot}
fit3 <- lm(log(mpg) ~ wt + factor(am), mtcars)
summary(fit3)$coef
ggplot(mtcars, aes(x= wt, y = log(mpg), colour= factor(am))) +
  geom_point(size = 4) + 
  geom_abline(intercept = summary(fit3)$coef[1,1], slope = summary(fit3)$coef[2,1], colour = "red") + 
  geom_abline(intercept = summary(fit3)$coef[1,1] + summary(fit3)$coef[3,1], slope = summary(fit3)$coef[2,1], colour = "blue") + 
  geom_hline(yintercept = mean(log(mtcars$mpg)[mtcars$am ==0]), colour = "red") +
  geom_hline(yintercept = mean(log(mtcars$mpg)[mtcars$am ==1]), colour = "blue")
```

I made a log trasformation to have linearity. It seems that am is highly related to weight: is car's weight is less than 3, you are very likely seing a manual trasmission, otherwise an automatic one. Expert knows that automatic trasmission cars weigh more. Now, weight is highly correlated to fule consumtpion. Does am tell us something more than weight? You see that adjusting for wt change order od magnitude and sign of the main effect (red horizontal line is mean of manual trasmission, blue is automatic)
; adjusting for weights probably make am a useless variable to predict mpg. Let's check with an ANOVA table

```{r am and wt}
fit0 <- lm(log(mpg) ~ wt, mtcars)
fit1 <- lm(log(mpg) ~ wt + factor(am), mtcars)
anova(fit0, fit1)
```

High p-value.
Let's interpret the last model:

```{r am and wt lm}
fit1 <- lm(log(mpg) ~ wt + factor(am), mtcars)
summary(fit1)
```

-0.04 is the change in the intercept if the trasmission move from manual to automatic : even a small decrease, while -0.28 is the percentage change in the mean of Y for a 1% change in the coefficient (for the log trasmormation). Let'see the confidence interval for these coefficient

```{r am wt confidence}
confint(fit1)
```

The change in the intercept could be also positive, but anyway we don't have overlapping points for any level f X, so we heavily rely on the model for comparning the behaviour of am on mpg for a certain level of X (there are no read point close to the Y axis.) I dont'know if this is because of the small sample or if because there are no automatic transmission car with those weights.

Let'see if this model violates some assumtpions:

```{r wt am diagnostic plot}
par(mfrow = c(2, 2))
plot(fit1)
```

TO BE CONTINUED


###Drat
```{r am drat}
ggplot(data = mtcars, aes(x = drat, y = mpg, colour = factor(am))) +
  geom_point(size = 5)
```

`drat` can explain some of the variance of `mpg` due to `am`, as you can see from the two groups at the side of the plot, but there are also some level of `drat` where `am` is not related to `mpg`

Let's see on an ANOVA table if adding `am` in a model with only `drat` as predictor is significant:

```{r am and drat}
fit0 <- lm(mpg ~ drat, mtcars)
fit1 <- lm(mpg ~ drat + am, mtcars)
anova(fit0, fit1)
```

Is not significant.
